LCS Tool: A Computational Platform for Lagrangian Coherent Structures

### 2014-10-06 main.tex
Update code listings to match J. Comp. Sci. submssion
MATLAB code listings updated to match Journal of Computational Science
submission by Florian Huhn on 2014-06-13.

\begin{table}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline
\multicolumn{2}{|p{\textwidth}|}
{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\tabularnewline
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\tabularnewline
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\tabularnewline
\hline
\lstinline!lambda! & scalar lambda value in \cref{eq:eta}\tabularnewline
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\tabularnewline
\hline
\lstinline!domain! & array to define flow domain\tabularnewline
\hline
\lstinline!resolution! & $1 \times 2$ array to define main grid resolution for Cauchy-Green strain tensor\tabularnewline
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\tabularnewline
\hline
\lstinline!PSList! & user-defined structure for Poincare section end-points, number of $\lambda$-lines launched from Poincare section, and maximum closed $\lambda$-line length\tabularnewline
\hline
\lstinline!nBisection! & optional number of bisection steps to refine zero crossings of Poincare map. Default: 5.\tabularnewline
\hline
\lstinline!dThresh! & optional threshold to discard discontinuous zero crossings of Poincare map. Default: $10^{-2}$.\tabularnewline
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify $\lambda$-line integration parameters\tabularnewline
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default: \lstinline![false,false]!.\tabularnewline
\hline
\end{tabular}
\caption{Syntax of LCS Tool's elliptic LCS functions.}
\label{t:Elliptic LCS functions}
\end{table}

The first step is to define the position of Poincare sections in regions where closed $\lambda$-lines are expected based on a visual analysis of the orbit structure of the $\eta_\pm^\lambda$ vector field. The Poincare section is to be oriented such that the first endpoint is close to the center of the expected Lagrangian vortex, and the second endpoint is outside this vortex. There is no automated procedure procedure implemented in LCS Tool for the positioning of Poincare sections (see however \citet{karrasch14:_autom_lagran} for a recently developed algorithm). Additionally, the number of lambda-lines launched from the Poincare section, \lstinline!poincareSection.numPoints!, must be defined. A reasonable default value is 100.

The second step is to integrate the $\lambda$-lines  starting from the Poincare section to obtain the corresponding Poincare map. Integration of $\lambda$-lines is performed using the $\eta_\pm^\lambda$ vector fields defined in \cref{eq:eta} over the main grid. The underlying eigenvector fields, $\xi_i(x_0)$, have generic but removable orientation discontinuities, which require monitoring and local reorientation. This process is sketched in \cref{t:variable step integration} and illustrated in \cref{f:variable step integration}. Linear interpolation is used in the interpolation of $\eta_\pm^\lambda$ within a grid element, since using higher-order interpolation would necessitate verifying that there are no orientation discontinuities beyond the four nearest grid points. We identify orientation discontinuities by checking the inner product of the $\eta_\pm^\lambda$ vectors at adjacent grid points. Rotations exceeding 90°, between two such neighboring vectors are classified as orientation discontinuities and are corrected before linear interpolation. When setting the Cauchy-Green strain tensor main grid resolution, one may find it helpful to calculate a histogram of eigenvector-field rotations to ensure that all rotations are well below 90° or almost 180°.

@@

\begin{figure}[hbt]
\centering
\includegraphics[width=.49\textwidth]{graphics/ocean_dataset/LCS_fwd_coherent_eddy}
\includegraphics[width=.49\textwidth]{graphics/ocean_dataset/LCS_fwd_colortracer}
\caption{(Left) Elliptic LCS (green) with repelling LCSs (red) continuing into its interior. (Right) Final positions $x(t) = (x_1(t),x_2(t)) = F_{t_0}^t(x_0)$ of tracer particles, are plotted with their $x_1(t)$ position encoded in red and $x_2(t)$ position encoded in green. Note the continued impact of the repelling hyperbolic LCSs (black) on tracer patterns inside the elliptic LCS (white).}
\label{f:ocean_dataset_colortracer}
\end{figure}

@@

In what follows, the parameter values are: $A = 0.1$, $\epsilon = 0.1$, $\omega = \pi/5$. The flow timespan is $t \in [0,10]$ and the domain is $x \in [0,2]$, $y \in [0,1]$. By examining the FTLE field (which
is optionally provided by LCS Tool), we position Poincare sections to capture elliptic LCSs. An LCS Tool demo script to perform this operation is given in \cref{l:double gyre lambda-line LCS} where two Poincare sections are defined. The free stretching parameter $\lambda$ (\cref{eq:etafields}) is varied over the range $[0.93,1.07]$ with increments of $0.01$. In a loop , closed orbits for all predefined $\lambda$ values are computed and the outermost closed orbit is kept as the Lagrangian vortex boundary. In this script, we have chosen small values for the error tolerances of the integration of the Cauchy-Green strain tensor and $\lambda$-lines to facilitate convergence.

\lstinputlisting[caption={Double gyre script for elliptic lambda-line LCSs},label=l:double gyre lambda-line LCS]{listings/double_gyre/lambda_lcs.m}

In \cref{f:double gyre lambda LCS convergence}, the resolution of the  Cauchy-Green strain tensor is varied from $500 \times 250$ to $1000 \times 500$. The location of the outermost closed lambda-line changes little, demonstrating convergence. For all tested resolutions the $\lambda$ values of the outermost closed orbits are constant. This is also suggests that the lowest tested resolution of $500 \times 250$ is sufficient to identify the elliptic LCSs in this example.

@@

\Cref{f:double gyre lambda hyperbolic LCS} shows the complete picture of elliptic and hyperbolic LCSs of the double gyre at this resolution. \Cref{lst:double_gyre_lambda_strain_stretch_lcs} lists the corresponding LCS Tool commands. The maximal length of shrinklines and stretchlines, \lstinline!strainlineMaxLength! and \lstinline!stretchlineMaxLength!, is set to $20$, a multiple of the domain size, since the hyperbolic LCSs may wind around vortices several times.

As usual, the local maximization distance is set larger for stretchlines than for shrinklines (cf. \cref{ll:strainlineLocalMaxDistance} and \cref{ll:stretchlineLocalMaxDistance}). The purpose of the maximization distance is to obtain spatially separated LCSs and to avoid a dense tangle of lines that basically indicates the same hyperbolic LCS (cf. \cref{t:Hyperbolic LCS algorithm}, \cref{i:flag maxima}). Setting the local maximization distance for stretchlines larger than for shrinklines allows obtaining a comparable number of stretchlines and shrinklines overall in the flow domain (recall that hyperbolic LCS seed points are discarded if they are within the local maximization distance of an existing hyperbolic LCS). Shrinklines are locally tangent to ridges of $\lambda_2$ maxima, whereas stretchlines are locally normal to these ridges. Setting the local maximization distance of stretchlines and shrinklines equal would therefore produce a greater number of stretchlines than shrinklines overall.

\lstinputlisting[
caption={LCS Tool commands for double gyre hyperbolic LCSs. Condensed version of the LCS Tool script \lstinline!demo/double_gyre/hyperbolic_shear_lcs.m!},
label=lst:double_gyre_lambda_strain_stretch_lcs
]
{listings/double_gyre/lambda_strain_stretch_lcs.m}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{graphics/double_gyre/lambda_hyperbolic_lcs}
\caption{LCSs in the double gyre flow. Resolution $[500\times250]$. $\lambda = 1.00$ and $1.04$ for the left and the right gyre. $\lambda \in [0.93,1.07],\,\Delta\lambda=0.01$. (Top) Elliptic (green) and repelling (red) LCSs with the FTLE shown in the background. (Bottom) Elliptic and attracting (blue) LCSs.}
\label{f:double gyre lambda hyperbolic LCS}
\end{figure}

@@

The integration time is $T=4L_x/U$, a multiple of a the eddy turnover time in the flow. \Cref{f:Bickley jet LCS} shows lambda-line LCS of the Bickley jet with the FTLE in the background. The figure demonstrates a poor result since the LCSs change shape significantly between eddies. To address this problem we could adjust numerical parameters such as the resolution, but instead we will analyze lambda-line closed orbits of one vortex for a short timespan.

\Cref{f:Bickley jet LCS} shows elliptic and hyperbolic LCSs of the Bickley jet with the FTLE field in the background.

\begin{figure}
\centering
\includegraphics[width=.85\textwidth]{graphics/bickley_jet/lambda_hyperbolic_lcs}
\caption{Elliptic and hyperbolic (top: red - repelling, bottom: blue - attracting) LCSs in the Bickley jet. $\lambda$-valuestrains of coherent eddies from left to right [0.95,\,0.80,\,0.94,\,0.80,\,0.94], $\lambda \in [0.80,1.20],\,\Delta\lambda=0.01$.}
\label{f:Bickley jet LCS}
\end{figure}

\lstinputlisting[caption={Bickley jet script for hyperbolic and elliptic LCSs. Subset of \lstinline!demo/bickley_jet/hyperbolic_shear_lcs.m!},label=lst:Bickley jet hyperbolic_shear_lcs subset]{listings/bickley_jet/hyperbolic_shear_lcs_subset.m}

\subsection{Ocean dataset}
\label{sec:oceandataset}

@@

The data is given at a spatial resolution of $1/4\degree$ and a temporal resolution of $7\,\mathrm{days}$. Due to the discrete data, defining the right hand side of \cref{eq:dtlon} and \cref{eq:dtlat} involves spline interpolation in space and time. An interpolant is generated first, then the function \lstinline!flowdata_derivative! evaluates the interpolants for the zonal and meridional velocity at the needed coordinates. \Cref{l:griddedInterp} shows the relevant part of the code in LCS Tool's ocean demo file \lstinline!hyperbolic_shear_lcs.m!. In \crefrange{ll:interpolation start}{ll:interpolation end}, the commands for the interpolation are given of the velocity dataset are given.

\begin{lstlisting}[caption={LCS ocean demo: Parts of the demo script included in LCS Tool to compute LCS from an ocean dataset},label=l:griddedInterp]
% Input parameters
domain = [0,6;-34,-28];
resolution = [400,400]; %@\label{ll:resolution}
timespan = [100,130]; %@\label{ll:timespan}

% Velocity definition
...
interpMethod = 'spline'; %@\label{ll:interpolation start}
vlon_interpolant = griddedInterpolant({time,lat,lon},vlon,interpMethod);
vlat_interpolant = griddedInterpolant({time,lat,lon},vlat,interpMethod);
lDerivative = @(t,x,~)flowdata_derivative(t,x,vlon_interpolant,vlat_interpolant); %@\label{ll:interpolation end}
incompressible = true; %@\label{ll:incompressible}

% LCS parameters
% Cauchy-Green strain
cgEigenvalueFromMainGrid = false; %@\label{ll:cgEigenvalueFromMainGrid}
cgAuxGridRelDelta = 0.01; %@\label{ll:cgAuxGridRelDelta}

% Lambda-lines
lambda = 1;
lambdaLineOdeSolverOptions = odeset('relTol',1e-6);

% Strainlines
strainlineMaxLength = 20;
gridSpace = diff(domain(1,:))/(double(resolution(1))-1);
strainlineLocalMaxDistance = 2*gridSpace;
strainlineOdeSolverOptions = odeset('relTol',1e-4);

% Stretchlines
stretchlineMaxLength = 20;
stretchlineLocalMaxDistance = 4*gridSpace;
stretchlineOdeSolverOptions = odeset('relTol',1e-4);
...
% Cauchy-Green strain eigenvalues and eigenvectors
[cgEigenvector,cgEigenvalue] = eig_cgStrain(lDerivative,domain,resolution,timespan,'incompressible',incompressible,'eigenvalueFromMainGrid',cgEigenvalueFromMainGrid,'auxGridRelDelta',cgAuxGridRelDelta); %@\label{ll:eig_cgStrain}

% Lambda-line LCSs
% Define Poincare sections; ...
poincareSection = struct('endPosition',{},'numPoints',{},'orbitMaxLength',{});
...
poincareSection(1).endPosition = [3.3,-32.1;3.7,-31.6]; %@\label{ll:poincareSection(1)}
poincareSection(2).endPosition = [1.3,-30.9;1.9,-31.1]; %@\label{ll:poincareSection(2)}

% Number of orbit seed points along each Poincare section
[poincareSection.numPoints] = deal(100);

% Set maximum orbit length to twice the expected circumference
nPoincareSection = numel(poincareSection);
for i = 1:nPoincareSection
    rOrbit = hypot(diff(poincareSection(i).endPosition(:,1)),diff(poincareSection(i).endPosition(:,2)));
    poincareSection(i).orbitMaxLength = 2*(2*pi*rOrbit);
end
[shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda); %@\label{ll:lambda_line}
closedLambdaLine = poincare_closed_orbit_multi(domain,resolution,shearline,poincareSection,'odeSolverOptions',lambdaLineOdeSolverOptions); %@\label{ll:poincare_closed_orbit_multi}
...
% Hyperbolic strainline LCSs
strainlineLcs = seed_curves_from_lambda_max(strainlineLocalMaxDistance,strainlineMaxLength,cgEigenvalue(:,2),cgEigenvector(:,1:2),domain,resolution,'odeSolverOptions',strainlineOdeSolverOptions); %@\label{ll:strainlineLcs}
...
stretchlineLcs = seed_curves_from_lambda_max(stretchlineLocalMaxDistance,stretchlineMaxLength,-cgEigenvalue(:,1),cgEigenvector(:,3:4),domain,resolution,'odeSolverOptions',stretchlineOdeSolverOptions); %@\label{ll:stretchlineLcs}
\end{lstlisting}

We choose the integration time as $T=30$ days (\cref{l:griddedInterp}, \cref{ll:timespan}), which is larger than the eddy turnover time in this region. The resolution of the main computational grid for initial conditions is set to $400 \times 400$ (\cref{ll:resolution}). This corresponds to a resolution of roughly 0.015° and gives good results. With this choice, the resolution of the tracer grid is 15 times higher than the resolution of the velocity field. The flow is integrated and the Cauchy-Green strain tensor is computed by the function \lstinline!eig_cgStrain! (\cref{ll:eig_cgStrain}). Incompressibility of the flow is enforced (\cref{ll:incompressible}), the auxiliary grid distance is set to 1\% of the main grid distance (\cref{ll:cgAuxGridRelDelta}), and eigenvalues are computed from the auxiliary grid (\cref{ll:cgEigenvalueFromMainGrid}). Elliptic LCSs are computed in \cref{ll:poincare_closed_orbit_multi}, after the Poincare sections have been set (\cref{ll:poincareSection(1),ll:poincareSection(2)}) and the $\eta_\pm^\lambda$ fields have been defined (\cref{ll:lambda_line}). Hyperbolic LCSs are computed in \cref{ll:strainlineLcs,ll:stretchlineLcs}.

>>>>

% FIXME Need to update table to account for lambda range
\begin{table}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline
\multicolumn{2}{|p{\textwidth}|}
{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\tabularnewline
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\tabularnewline
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\tabularnewline
\hline
\lstinline!lambda! & scalar lambda value in \cref{eq:eta}\tabularnewline
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\tabularnewline
\hline
\lstinline!domain! & array to define flow domain\tabularnewline
\hline
\lstinline!resolution! & $1 \times 2$ array to define main grid resolution for Cauchy-Green strain tensor\tabularnewline
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\tabularnewline
\hline
\lstinline!PSList! & user-defined structure for Poincare section end-points, number of $\lambda$-lines launched from Poincare section, and maximum closed $\lambda$-line length\tabularnewline
\hline
\lstinline!nBisection! & optional number of bisection steps to refine zero crossings of Poincare map. Default: 5.\tabularnewline
\hline
\lstinline!dThresh! & optional threshold to discard discontinuous zero crossings of Poincare map. Default: $10^{-2}$.\tabularnewline
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify $\lambda$-line integration parameters\tabularnewline
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default: \lstinline![false,false]!.\tabularnewline
\hline
\end{tabular}
\caption{Syntax for LCS Tool elliptic LCS functions.}
\label{t:Elliptic LCS functions}
\end{table}

The first step is to define the position of Poincare sections in regions where closed $\lambda$-lines are expected based on a visual analysis of the orbit structure of the $\eta_\pm^\lambda$ vector field. The Poincare section is to be oriented such that the first endpoint is close to the center of the expected Lagrangian vortex, and the second endpoint is outside this vortex. There is no automated procedure implemented in LCS Tool for the positioning of Poincare sections (see however \citet{karrasch14:_autom_lagran} for a recently developed algorithm). Additionally, the number of lambda-lines launched from the Poincare section, \lstinline!poincareSection.numPoints!, must be defined. A reasonable default value is 100.

The second step is to integrate the $\lambda$-lines  starting from the Poincare section to obtain the corresponding Poincare map. Integration of $\lambda$-lines is performed using the $\eta_\pm^\lambda$ vector fields defined in \cref{eq:eta} over the main grid. The underlying eigenvector fields, $\xi_i(x_0)$, have generic but removable orientation discontinuities, which require monitoring and local reorientation. This process is sketched in \cref{t:variable step integration} and illustrated in \cref{f:variable step integration}. Linear interpolation is used in the interpolation of $\eta_\pm^\lambda$ within a grid element, since using higher-order interpolation would necessitate verifying that there are no orientation discontinuities beyond the four nearest grid points. We identify orientation discontinuities by checking the inner product of the $\eta_\pm^\lambda$ vectors at adjacent grid points. Rotations exceeding 90°, between two such neighboring vectors are classified as orientation discontinuities and are corrected before linear interpolation. When setting the Cauchy-Green strain tensor main grid resolution, one may find it helpful to calculate a histogram of eigenvector field rotations to ensure that all rotations are well below 90° or almost 180°.

@@

\begin{figure}[hbt]
\centering
\includegraphics[width=.49\textwidth]{graphics/ocean_dataset/LCS_fwd_coherent_eddy}
\includegraphics[width=.49\textwidth]{graphics/ocean_dataset/LCS_fwd_colortracer}
\caption{(Left) Elliptic LCS (green) with repelling LCSs (red) continuing into its interior. (Right) Final positions $x(t) = (x_1(t),x_2(t)) = F_{t_0}^t(x_0)$ of tracer particles. $x_1(t)$ position encoded in red and $x_2(t)$ position encoded in green. Note the continued impact of the repelling hyperbolic LCSs (black) on tracer patterns inside the elliptic LCS (white).}
\label{f:ocean_dataset_colortracer}
\end{figure}

@@

In what follows, the parameter values are: $A = 0.1$, $\epsilon = 0.1$, $\omega = \pi/5$. The flow timespan is $t \in [0,10]$ and the domain is $x \in [0,2]$, $y \in [0,1]$. By examining the FTLE field (which LCS Tool can provide), we position Poincare sections to capture elliptic LCSs. An LCS Tool demo script to perform this operation is given in \cref{l:double gyre lambda-line LCS} where two Poincare sections are defined (\cref{ll:double gyre poincareSection(1),ll:double gyre poincareSection(2)}). The free stretching parameter $\lambda$ (\cref{eq:etafields}) is varied over the range $[0.93,1.07]$ with increments of $0.01$ (\cref{ll:double gyre lambdaRange}).
% FIXME The cited loop contains ellipses (...), that hide important code and should become an LCS Tool function
In a loop (\crefrange{ll:double gyre lambda loop start}{ll:double gyre lambda loop end}), closed orbits for all predefined $\lambda$ values are computed and the outermost closed orbit is kept as the Lagrangian vortex boundary.
% FIXME Small tolerances (i.e. strict tolerances) should make convergence difficult not easy, so the word "facilitate" in the sentence below seems illogical.
In this script, we have chosen small values for the error tolerances of the integration of the Cauchy-Green strain tensor (\cref{ll:double gyre cgStrainOdeSolverOptions}) and $\lambda$-lines (\cref{ll:double gyre lambdaLineOdeSolverOptions})to facilitate convergence.

\lstinputlisting[caption={LCS Tool commands for double gyre elliptic LCSs. Subset of the LCS Tool script \lstinline!demo/double_gyre/hyperbolic_shear_lcs.m!},
                 label=l:double gyre lambda-line LCS]{listings/double_gyre/elliptic_lcs_subset.m}

In \cref{f:double gyre lambda LCS convergence}, the resolution of the  Cauchy-Green strain tensor is varied from $500 \times 250$ to $1000 \times 500$. The location of the outermost closed $\lambda$-line changes insignificantly, demonstrating convergence. For all tested resolutions, the $\lambda$ values of the outermost closed orbits match.
% FIXME Should reduce lowest resolution used to demonstrate problems if selected resolution is too low
This also suggests that the lowest tested resolution, $500 \times 250$, is sufficient to identify elliptic LCSs.

@@

\Cref{f:double gyre lambda hyperbolic LCS} shows the complete picture of elliptic and hyperbolic LCSs of the double gyre at this resolution. \Cref{l:double gyre lambda strain stretch LCS} lists the corresponding LCS Tool commands. The maximal length of shrinklines and stretchlines, \lstinline!strainlineMaxLength! and \lstinline!stretchlineMaxLength!, is set to $20$, a multiple of the domain size, since the hyperbolic LCSs may wind around vortices several times.

As usual, the local maximization distance is set larger for stretchlines than for shrinklines (cf. \cref{ll:shrinklineLocalMaxDistance} and \cref{ll:stretchlineLocalMaxDistance}). The purpose of the maximization distance is to obtain spatially separated LCSs and to avoid a dense tangle of lines that basically indicates the same hyperbolic LCS (cf. \cref{t:Hyperbolic LCS algorithm}, \cref{i:flag maxima}). Setting the local maximization distance for stretchlines larger than for shrinklines allows obtaining a comparable number of stretchlines and shrinklines overall in the flow domain (recall that hyperbolic LCS seed points are discarded if they are within the local maximization distance of an existing hyperbolic LCS). Shrinklines are locally tangent to ridges of $\lambda_2$ maxima, whereas stretchlines are locally normal to these ridges. Setting the local maximization distance of stretchlines and shrinklines equal would therefore produce a greater number of stretchlines than shrinklines overall.

\lstinputlisting[
        caption={LCS Tool commands for double gyre hyperbolic LCSs. Subset of the LCS Tool script \lstinline!demo/double_gyre/hyperbolic_shear_lcs.m!},
        label=l:double gyre lambda strain stretch LCS]
        {listings/double_gyre/hyperbolic_lcs_subset.m}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{graphics/double_gyre/lambda_hyperbolic_lcs}
\caption{LCSs in the double gyre flow. Resolution $[500\times250]$. $\lambda = 1.00$ and $1.04$ for the left and the right gyre. $\lambda \in [0.93,1.07]$, $\Delta\lambda = 0.01$.
(Top) Elliptic (green) and repelling (red) LCSs with the FTLE shown in the background. (Bottom) Elliptic and attracting (blue) LCSs.}
\label{f:double gyre lambda hyperbolic LCS}
\end{figure}

@@

The integration time is $T=4L_x/U$, a multiple of a the eddy turnover time in the flow. \Cref{l:Bickley jet elliptic hyperbolic LCS subset} shows the LCS Tool commands in which the chaotically perturbed velocity is defined (\crefrange{ll:Bickley jet velocity definition start}{ll:Bickley jet velocity definition end}), periodic boundary conditions are imposed in the x-direction (\cref{ll:Bickley jet periodicBc}), and five Poincare sections are defined where we expect coherent vortices (\crefrange{ll:Bickley jet Poincare start}{ll:Bickley jet Poincare end}).
$\lambda$-values for closed orbit detection are varied over the range $[0.80,1.20]$ with a step of $0.01$.

\Cref{f:Bickley jet LCS} shows elliptic and hyperbolic LCSs of the Bickley jet with the FTLE in the background.

\lstinputlisting[caption={LCS Tool commands for Bickley jet elliptic and hyperbolic LCSs. Subset of the LCS Tool script \lstinline!demo/bickley_jet/hyperbolic_shear_lcs.m!},
                 label=l:Bickley jet elliptic hyperbolic LCS subset]{listings/bickley_jet/elliptic_hyperbolic_lcs_subset.m}

\begin{figure}
\centering
\includegraphics[width=.85\textwidth]{graphics/bickley_jet/lambda_hyperbolic_lcs}
\caption{Elliptic and hyperbolic (top: red - repelling, bottom: blue - attracting) LCSs in the Bickley jet. $\lambda$ values of elliptic LCSs from left to right are [0.95,\,0.80,\,0.94,\,0.80,\,0.94], $\lambda \in [0.80,1.20]$, $\Delta\lambda = 0.01$.}
\label{f:Bickley jet LCS}
\end{figure}

\subsection{Ocean velocity data from satellite altimetry}
\label{sec:oceandataset}

@@

The data is given at a spatial resolution of $1/4\degree$ and a temporal resolution of 7 days.
Due to the discrete data, defining the right hand side of \cref{eq:dtlon} and \cref{eq:dtlat} involves spline interpolation in space and time.
An interpolant is generated first, then the function \lstinline!flowdata_derivative! evaluates the interpolants for the zonal and meridional velocity at the needed coordinates.
\Cref{l:ocean data elliptic hyperbolic lcs} shows the relevant part of the code in LCS Tool's ocean demo file \lstinline!demo/ocean_dataset/hyperbolic_shear_lcs.m!.
The commands for the interpolation of the velocity data are given in \crefrange{ll:interpolation start}{ll:interpolation end}.

\lstinputlisting[
        caption={LCS Tool commands for ocean data elliptic and hyperbolic LCSs. Subset of the LCS Tool script \lstinline!demo/ocean_dataset/hyperbolic_shear_lcs.m!},
        label={l:ocean data elliptic hyperbolic lcs}]
        {listings/ocean_dataset/elliptic_hyperbolic_lcs_subset.m}

We choose the integration time as $T=30$ days (\cref{l:ocean data elliptic hyperbolic lcs}, \cref{ll:ocean data timespan}), which is larger than the eddy turnover time in this region. The resolution of the main computational grid for initial conditions is set to $400 \times 400$ (\cref{ll:ocean data resolution}). This corresponds to a resolution of roughly 0.015° and gives good results. With this choice, the resolution of the tracer grid is 15 times higher than the resolution of the velocity field. The flow is integrated and the Cauchy-Green strain tensor is computed by the function \lstinline!eig_cgStrain! (\cref{ll:eig_cgStrain}). Incompressibility of the flow is enforced (\cref{ll:incompressible}), the auxiliary grid distance is set to 1\% of the main grid distance (\cref{ll:cgAuxGridRelDelta}), and eigenvalues are computed from the auxiliary grid (\cref{ll:cgEigenvalueFromMainGrid}). Elliptic LCSs are computed in \cref{ll:ocean data poincare_closed_orbit_multi}, after the Poincare sections have been set (\cref{ll:ocean data poincareSection(1),ll:ocean data poincareSection(2)}) and the $\eta_\pm^\lambda$ fields have been defined (\cref{ll:ocean data lambda_line}). $\lambda$-values are varied over a range of $[0.90,1.10]$ with a step of $0.02$ (\cref{ll:ocean data lambdaRange}).
Hyperbolic LCSs are computed in \cref{ll:ocean data shrinklineLcs,ll:ocean data stretchlineLcs}.
